### 2025-01-06 sections/2-methods.tex
some more cleanup

Given a dataset $\textbf{X}$, a distance function $f$ and the root cluster of the tree built using Algorithm~\ref{alg:methods:partition}, we can now formally pose the $k$-NN search problem.

The $k$-NN search problem, given a user specified integer $k$ and query $q$, aims to find the $k$ closest points to $q$ in $\textbf{X}$.
In other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.

In this section, we present three novel algorithms for exact $k$-NN search:
Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.

In these algorithms, we use $H$, for \textit{hits}, to refer to the data structure that stores the closest points to the query found so far, and $Q$ to refer to the data structure that stores the clusters and points that are still in contention for being one of the $k$ nearest neighbors.
The two Sieve algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

@@

For Repeated $\rho$-NN search, we begin by performing \textit{tree-search} with a search radius $r$ equal to the radius of the root cluster divided by the cardinality of the dataset.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.

@@

Once $\sum_{C \in Q} |C| \geq k$, we have found at least $k$ neighbors, and so we can stop increasing the radius.
We now perform \textit{leaf-search} to find the $k$ nearest neighbors among the points in the clusters found during the previous steps.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

        \State $r \Leftarrow radius$ of the $root$ cluster
        \State $r \Leftarrow$ $\frac{r}{|root|}$

        \Loop
            \State $Q \Leftarrow$ tree-search($root$, $q$, $r$)
            \If{$Q \neq \emptyset$}
                \State \textbf{break}
            \EndIf
            \State $r \Leftarrow 2 \cdot r$
        \EndLoop

        \Loop
            \If{$\sum_{C \in Q} |C| >= k$}
                \State \textbf{break}
            \EndIf
            \State $\mu \Leftarrow |Q| \cdot \left( \sum_{C \in Q} LFD(C)^{-1} \right)^{-1}$
            \State $r \Leftarrow r \cdot \min \left( 2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu^{-1}} \right)$
            \State $Q \Leftarrow$ tree-search($root$, $q$, $r$)
        \EndLoop

        \State $H \Leftarrow \text{leaf-search}(Q, q, r)$
        \State \textbf{sort} $H$ by non-decreasing distance to $q$
        \State \textbf{return} The first $k$ points in $H$
    \end{algorithmic}
\end{algorithm}

@@

Based on the assumption that the LFD near the query does not differ significantly from that of the clusters near the query (the set $Q$), Equation~\ref{eq:methods:repeated-rnn-factor} suggests that in the expected case, we need only two iterations of tree-search to find $k$ neighbors:
one iteration to find at least one cluster, and one more to find enough of the $k$ neighbors.
Since this is a constant factor, complexity of tree-search for Repeated $\rho$-NN is the same as that of $\rho$-NN search, i.e.,\,$\mathcal{O}\big(\log\mathcal{N}_{\hat{r}}(X)\big)$.

@@

\begin{minipage}{0.5125\textwidth}
    \begin{algorithm}[H]\small
        \caption{Breadth-First Sieve($root$, $q$, $k$)} % give the algorithm a caption
        \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
        \begin{algorithmic} % enter the algorithmic environment
            \Require $root$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find

            \State $c \Leftarrow$ \textit{center} of $root$
            \State $Q \Leftarrow$ \{ ($root$, $\delta^{+}_{root}$, $|root| - 1$), ($c$, $\delta_{root}$, 1) \}

            \Loop
                \If{$k = \sum_{(\_, \_, m) \in Q} m$}
                    \State \textbf{break}
                \EndIf
                \State $\tau \Leftarrow$ QuickSelect($Q$, $k$)
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \If{$\delta^{-}_{p} > \tau$}
                        \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \EndIf
                \EndFor
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \If{$p$ is a point}
                        \State \textbf{continue}
                    \ElsIf{$p$ is a leaf}
                        \For{$c \in p$}
                            \State $Q \Leftarrow Q \cup \{ (c, \delta_{c}, 1) \}$
                        \EndFor
                    \Else
                        \State $[L, R] \Leftarrow$ children of $p$
                        \State $Q \Leftarrow Q \cup \{ (L, \delta^{+}_{L}, |L| - 1), (L_c, \delta_{L}, 1) \}$
                        \State $Q \Leftarrow Q \cup \{ (R, \delta^{+}_{R}, |R| - 1), (R_c, \delta_{R}, 1) \}$
                    \EndIf
                \EndFor
            \EndLoop

            \State \textbf{return} $Q$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{0.425\textwidth}
    \begin{algorithm}[H]\small
        \caption{Depth-First Sieve($root$, $q$, $k$)}
        \label{alg:methods:depth-first-sieve}
        \begin{algorithmic}
            \Require $root$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find

@@

            \While{$|H| < k$ \textbf{or} $H.peek.\delta \geq Q.peek.\delta^{-}$}
                \While{$\neg (Q.peek$ is a leaf)}
                    \State{$C \Leftarrow Q.pop$, the closest cluster}
                    \State{$[L, R] \Leftarrow$ children of $C$}
                    \State{$Q.push(L)$}
                    \State{$Q.push(R)$}
                \EndWhile
                \State{$leaf \Leftarrow Q.pop$}
                \For{$p \in leaf$}
                    \State{$H.push(p)$}
                \EndFor
                \While{$|H| > k$}
                    \State{$H.pop$}
                \EndWhile
            \EndWhile
            \State Return $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

@@

This algorithm performs a breadth-first traversal of the tree, pruning clusters by using a modified version of the QuickSelect algorithm~\cite{hoare1961algorithm} at each level.

@@

We then use the QuickSelect algorithm, modified to account for multiplicities and to reorder $Q$ in-place, to find the element in $Q$ with the $k^{th}$ smallest $\delta^{+}$; in other words, we find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B_X(q, \tau) \right| \geq k$.
Since this step may require a binary search for the correct pivot element to find $\tau$ and reordering with a new pivot takes linear time in the size of the input list, this version of QuickSelect has $\mathcal{O}(|Q| \log |Q|)$ time complexity.

@@

We continue this process until $Q$ no longer contains any clusters.
We then use the QuickSelect algorithm one last time to reorder $Q$, find $\tau$, and return the $k$ nearest neighbors.

This process is described in Algorithm~\ref{alg:methods:bredth-first-sieve}.

@@

This algorithm performs a depth-first traversal of the tree and uses two priority queues to track clusters and hits.

@@

This process is described in Algorithm~\ref{alg:methods:depth-first-sieve}.
It terminates when $H$ is full and the top priority element in $H$ has $\delta$ less than the top priority element in $Q$, i.e.,\,the theoretically closest point left to be considered in $Q$ is farther from the query than the $k^{th}$ nearest neighbor in $H$.
This leaves $H$ containing exactly the $k$ nearest neighbors to the query.

@@

Due to their similarity, we combine the complexity analyses of both Sieve methods.
For these methods we again use the terminology of tree-search and leaf-search.
Tree-search navigates the cluster tree and adds clusters to $Q$.
Leaf-search exhaustively searches some of the clusters in $Q$ to find the $k$ nearest neighbors.

>>>>

Given a dataset $\textbf{X}$, a distance function $f$ and the root cluster of the tree built using Algorithm~\ref{alg:methods:partition}, we can now formally pose the $k$-NN search problem.
Given a user specified integer $k$ and query $q$, $k$-NN search aims to find the $k$ closest points to $q$ in $\textbf{X}$.
In other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.

In this section, we present three novel algorithms for exact $k$-NN search:
Repeated $\rho$-NN, Breadth-First Sieve, and Depth-First Sieve.
In these algorithms, we use $H$, for \textit{hits}, to refer to the data structure that stores the closest points to the query found so far, and $Q$ to refer to the data structure that stores the clusters and points that are still in contention for being one of the $k$ nearest neighbors.
The two Sieve algorithms also use some terminology defined and illustrated in Figure~\ref{fig:methods:deltas}.

@@

For Repeated $\rho$-NN search (Algorithm~\ref{alg:methods:repeated-rnn}), we begin by performing \textit{tree-search} with a search radius $r$ equal to the radius of the root cluster divided by the cardinality of the dataset.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.

@@

Once $\sum_{C \in Q} |C| \geq k$, we are guaranteed to have found at least $k$ neighbors, and so we perform \textit{leaf-search} to find the $k$ nearest neighbors among the points in the clusters found after the last tree-search.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find
        \State $r \Leftarrow radius$ of the $root$ cluster
        \State $r \Leftarrow$ $\frac{r}{|root|}$
        \Loop
            \State $Q \Leftarrow$ tree-search($root$, $q$, $r$)
            \If{$Q \neq \emptyset$}
                \State \textbf{break}
            \EndIf
            \State $r \Leftarrow 2 \cdot r$
        \EndLoop
        \Loop
            \If{$\sum_{C \in Q} |C| >= k$}
                \State \textbf{break}
            \EndIf
            \State $\mu \Leftarrow |Q| \cdot \left( \sum_{C \in Q} LFD(C)^{-1} \right)^{-1}$
            \State $r \Leftarrow r \cdot \min \left( 2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu^{-1}} \right)$
            \State $Q \Leftarrow$ tree-search($root$, $q$, $r$)
        \EndLoop
        \State $H \Leftarrow \text{leaf-search}(Q, q, r)$
        \State \textbf{sort} $H$ by non-decreasing distance to $q$
        \State \textbf{return} The first $k$ points in $H$
    \end{algorithmic}
\end{algorithm}

@@

Based on the assumption that the LFD near the query does not differ significantly from that of the clusters near the query (the set $Q$), Equation~\ref{eq:methods:repeated-rnn-factor} suggests that in the expected case, we need only two iterations of tree-search to find $k$ neighbors:
one iteration to find at least one cluster, and one more to find enough clusters to guarantee $k$ neighbors.
Since this is a constant factor, complexity of tree-search for Repeated $\rho$-NN is the same as that of $\rho$-NN search, i.e.,\,$\mathcal{O}\big(\log\mathcal{N}_{\hat{r}}(X)\big)$.

@@

\begin{minipage}{0.5125\textwidth}
    \begin{algorithm}[H]\small
        \caption{Breadth-First Sieve($root$, $q$, $k$)} % give the algorithm a caption
        \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
        \begin{algorithmic} % enter the algorithmic environment
            \Require $root$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find
            \State $c \Leftarrow$ \textit{center} of $root$
            \State $Q \Leftarrow$ \{ ($root$, $\delta^{+}_{root}$, $|root| - 1$), ($c$, $\delta_{root}$, 1) \}
            \Loop
                \If{$k = \sum_{(\_, \_, m) \in Q} m$}
                    \State \textbf{break}
                \EndIf
                \State $\tau \Leftarrow$ QuickSelect($Q$, $k$)
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \If{$\delta^{-}_{p} > \tau$}
                        \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \EndIf
                \EndFor
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \If{$p$ is a point}
                        \State \textbf{continue}
                    \ElsIf{$p$ is a leaf}
                        \For{$c \in p$}
                            \State $Q \Leftarrow Q \cup \{ (c, \delta_{c}, 1) \}$
                        \EndFor
                    \Else
                        \State $[L, R] \Leftarrow$ children of $p$
                        \State $l, r \Leftarrow$ centers of $L, R$
                        \State $Q \Leftarrow Q \cup \{ (L, \delta^{+}_{L}, |L| - 1), (l, \delta_{L}, 1) \}$
                        \State $Q \Leftarrow Q \cup \{ (R, \delta^{+}_{R}, |R| - 1), (l, \delta_{R}, 1) \}$
                    \EndIf
                \EndFor
            \EndLoop
            \State QuickSelect($Q$, $k$)
            \State \textbf{return} The first $k$ points in $Q$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{0.425\textwidth}
    \begin{algorithm}[H]\small
        \caption{Depth-First Sieve($root$, $q$, $k$)}
        \label{alg:methods:depth-first-sieve}
        \begin{algorithmic}
            \Require $root$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find

@@

            \While{$|H| < k$ \textbf{or} $H.peek.\delta \geq Q.peek.\delta^{-}$}
                \While{$\neg (Q.peek$ is a leaf)}
                    \State{$C \Leftarrow Q.pop$, the closest cluster}
                    \State{$[L, R] \Leftarrow$ children of $C$}
                    \State{$Q.push(L)$}
                    \State{$Q.push(R)$}
                \EndWhile
                \State{$leaf \Leftarrow Q.pop$}
                \For{$p \in leaf$}
                    \State{$H.push(p)$}
                \EndFor
                \While{$|H| > k$}
                    \State{$H.pop$}
                \EndWhile
            \EndWhile
            \State \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

@@

This algorithm (described in Algorithm~\ref{alg:methods:bredth-first-sieve}) performs a breadth-first traversal of the tree, pruning clusters by using a modified version of the QuickSelect algorithm~\cite{hoare1961algorithm} at each level.

@@

We then use the QuickSelect algorithm, modified to account for multiplicities and to reorder $Q$ in-place, to find the element in $Q$ with the $k^{th}$ smallest $\delta^{+}$; in other words, we find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B(q, \tau) \right| \geq k$.
Since this step may require a binary search for the correct pivot element to find $\tau$ and reordering with a new pivot takes linear time in the size of the input list, this version of QuickSelect has $\mathcal{O}\left(|Q| \log |Q|\right)$ time complexity.

@@

We continue this process until $Q$ no longer contains any clusters.
We then use the QuickSelect algorithm one last time to reorder $Q$ and return the $k$ nearest neighbors.

@@

This algorithm (described in Algorithm~\ref{alg:methods:depth-first-sieve}) is quite similar to a depth-first traversal of the tree, using two priority queues to track clusters and hits, and to prioritize which branch of the tree to explore next.

@@

This process terminates when $H$ is full and the top priority element in $H$ has $\delta$ less than the top priority element in $Q$, i.e.,\,the theoretically closest point left to be considered in $Q$ is farther from the query than the $k^{th}$ nearest neighbor in $H$.
This leaves $H$ containing exactly the $k$ nearest neighbors to the query.

@@

Due to their similarity, we combine the complexity analyses of both Sieve methods.
For these methods we again use the terminology of tree-search and leaf-search.
Tree-search navigates the cluster tree and adds clusters to $Q$.
Leaf-search exhaustively searches \textit{some} of the clusters in $Q$ to find the $k$ nearest neighbors.

### 2025-01-06 sections/3-datasets.tex
moved some floats for better whitespace usage
some more cleanup

We benchmark on a variety of datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
Table~\ref{tab:datasets:summary} summarizes these datasets.
The benchmarks for Fashion-Mnist, Glove-25, Sift, and Random (and for their synthetic augmentations) were conducted on an Amazon AWS (EC2) \texttt{r6i.16xlarge} instance, with a 64-core Intel Xeon Platinum 8375C CPU 2.90GHz processor, 512GB RAM.
This is the same configuration used in~\cite{aumuller2020ann}.
The OS kernel was Ubuntu 22.04.3-Ubuntu SMP.
The Rust compiler was Rust 1.72.0, and the Python interpreter version was 3.9.16.
The benchmarks for SILVA and RadioML were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM.
The OS kernel was Manjaro Linux 5.15.130-1-MANJARO.
The Rust compiler was Rust 1.72.0, and the Python interpreter version was 3.9.16.

\begin{table}[!t]
    % \renewcommand{\arraystretch}{1.15}
    \caption{Datasets used in benchmarks.}
    \label{tab:datasets:summary}
    \vskip 0.15in
    \begin{center}
        \begin{small}
            \begin{sc}
                \begin{tabular}{|l|l|l|l|}
                    \hline
                    \textbf{Dataset} & \textbf{Distance}  &\textbf{Cardinality}  & \textbf{Dimensionality}  \\
                    \hline
                    Fashion-Mnist    & Euclidean              & 60,000             & 784                    \\
                    \hline
                    Glove-25         & Cosine                 & 1,183,514          & 25                     \\
                    \hline
                    Sift             & Euclidean              & 1,000,000          & 128                    \\
                    \hline
                    Random           & Euclidean              & 1,000,000          & 128                    \\
                    \hline
                    SILVA            & Levenshtein            & 2,224,640          & 3,712 - 50,000         \\
                    \hline
                    RadioML          & Dynamic Time Warping   & 97,920             & 1,024                  \\
                    \hline
                \end{tabular}
            \end{sc}
        \end{small}
    \end{center}
    \vskip -0.1in
\end{table}

\subsection{Random Datasets and Synthetic Augmentations}
\label{sec:datasets-and-benchmarks:random-datasets}

In addition to benchmarks on datasets from the ANN-Benchmarks suite, we also benchmarked on synthetic augmentations of these real datasets, using the process described in \ref{sec:methods:synthetic-data}.
In particular, we use a noise tolerance $\epsilon = 0.01$ and explore the scaling behavior as the cardinality multiplier (referred to as ``Multiplier'' in Table~\ref{tab:results:qps-and-recall}) increases.
We also benchmarked on purely randomly-generated datasets of various cardinalities.
For this, we used a base cardinality of 1,000,000 and a dimensionality of 128 to match the Sift dataset; hereafter, we refer to the random dataset with cardinality 1,000,000 and dimensionality 128 as ``Random.''
This benchmark allows us to isolate the effect of a manifold structure (which we expect to be absent in a purely random dataset) on the performance of the CAKES' algorithms.

>>>>

We benchmark on a variety of datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
Table~\ref{tab:datasets:summary} summarizes these datasets.
All benchmarks were conducted on an Intel Xeon E5-2690 v4 CPU @ 2.60GHz with 512GB RAM.
The OS kernel was Manjaro Linux 5.15.164-1-MANJARO.
The Rust compiler was Rust 1.83.0, and the Python interpreter version was 3.9.18.

\begin{table}
    \caption{Datasets used in benchmarks.}
    \label{tab:datasets:summary}
    \begin{center}
        \begin{sc}
            \begin{tabular}{|l|l|l|l|}
                \hline
                \textbf{Dataset} & \textbf{Distance Function}  &\textbf{Cardinality}  & \textbf{Dimensionality}  \\
                \hline
                Fashion-Mnist    & Euclidean                   & 60,000             & 784                    \\
                \hline
                Glove-25         & Cosine                      & 1,183,514          & 25                     \\
                \hline
                Sift             & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                Random           & Euclidean                   & 1,000,000          & 128                    \\
                \hline
                SILVA            & Levenshtein                 & 2,224,640          & 3,712 - 50,000         \\
                \hline
                RadioML          & Dynamic Time Warping        & 97,920             & 1,024                  \\
                \hline
            \end{tabular}
        \end{sc}
    \end{center}
    \vskip -0.1in
\end{table}

\subsection{Random Datasets and Synthetic Augmentation}
\label{sec:datasets-and-benchmarks:random-datasets}

In addition to benchmarks on datasets from the ANN-Benchmarks suite, we also benchmarked on synthetic augmentations of these real datasets, using the process described in \ref{sec:methods:synthetic-data}.
In particular, we use a noise tolerance $\epsilon = 0.01$ and explore the scaling behavior as the cardinality multiplier (referred to as ``Mult.'' in Tables~\ref{tab:results:qps-and-recall-fmn},~\ref{tab:results:qps-and-recall-glove},~\ref{tab:results:qps-and-recall-sift} and~\ref{tab:results:qps-and-recall-random}) increases.
We also benchmarked on purely randomly-generated datasets of various cardinalities.
For this, we used a base cardinality of 1,000,000 and a dimensionality of 128 to match the Sift dataset; hereafter, we refer to the random dataset with cardinality 1,000,000 and dimensionality 128 as ``Random.''
This benchmark allows us to isolate the effect of a manifold structure (which we expect to be absent in a purely random dataset) on the performance of the CAKES' algorithms.
